Neural Models for Target-Based Computer-Assisted Musical Orchestration: A Preliminary Study

### 2022-03-26 csmc_mume_2020/paper/JCMS_2021/paper.tex
JCMS paper camera ready edits

\usepackage{amssymb}
\usepackage{amsmath}
\setcounter{tocdepth}{3}
\usepackage{graphicx}
\usepackage{caption}
\usepackage{subcaption}
\captionsetup{labelfont=bf, labelsep=period,compatibility=false}
\usepackage{apacite}
\newcommand{\keywords}[1]{\par\addvspace\baselineskip
\noindent\keywordname\enspace\ignorespaces#1}

@@

\sloppypar{The codebase for this paper can be found at \url{https://github.com/dzluke/DeepOrchestration} and you can listen to audio examples at \url{https://dzluke.github.io/DeepOrchestration/}.}

@@

\begin{enumerate}
\item we train specific models to classify the instruments present in combinations of sounds from a database of instrument notes, up to ten simultaneous instruments;
\item we then pick the best classifier and we feed into it an unknown sound to be classified;
\item since the output of the classifier will be in the form of the probability that specific instruments-note pairs are present in the sound, we use this information to synthesize an orchestration for the target sound;
\item finally, we evaluate the generated orchestration against state-of-the-art systems for computer-assisted orchestrations.
\end{enumerate}

@@

Two neural classifiers have been trained on the problem of recognizing instrument-pitch pairs, which has a total of 424 classes (i.e. 424 unique instrument-pitch pairs). We want to point out, however, that this is still a simplified problem compared to assisted orchestration. A complete orchestration solution would normally be made by triples of instrument-pitch-dynamics. It is difficult to frame this problem as classification, since we would need to have a very high number of classes. Moreover, for the nature of the samples we use (a typical sample is in the form \texttt{Flute-C4-pp}, as described in section \ref{sec:dataset}), each class would be represented by a single sample. For these reasons, our models have been trained on simplified versions of the original problem, and we employed a different strategy to determine dynamics. When a network is given a target sound as input, it will attempt to apply the same classification rules, outputting a vector of the probabilities of each sample being in the target sound. Since the classes only encode instrument-pitch pairs but we also need the dynamics, we quantize the probability function generated by the network into three echelons coresponding, respectively, to the dynamic levels \emph{pianissimo}, \emph{mezzoforte} and \emph{fortissimo} (see section \ref{sec:orchestration}). By taking the instruments and pitches that have the highest probability and by using quantized probabilities to retrieve dynamics, a full orchestrated solution can be created.  Figure \ref{fig:method} illustrates the described ideas.

@@

To create the input data for training the classifiers, we used the \emph{TinySOL} database. TinySOL is a subset of the Studio On Line (SOL) database created by IRCAM \cite{Cella2020b}. TinySOL contains 1,529 samples from 12 instruments. The instruments come from different orchestral families: strings, woodwinds, and brass. Each sample is one instrument playing a single note in the \emph{ordinario} playing style, with one of three dynamics: \textit{pp}, \textit{mf}, or \textit{ff} (for example \texttt{Flute-C4-pp} or \texttt{Clarinet-D5-mf}). The instruments and ranges over which they were recorded are summarized in table \ref{tab:ranges}.

For a given number of instruments $N$, each input to our model is a combination of $N$ randomly selected TinySOL samples chosen from an orchestra of 10 instruments. The only constraint imposed on the random selection is that there cannot be more than 3 samples from the same instrument in any one combination. We do this to ensure instrumental variety in the combinations. By randomly selecting the samples, we achieve a variety of instruments, pitches, and dynamics while avoiding introducing any bias in the combinations that could result from hand-picking the samples. The selected samples are combined to be played simultaneously and the magnitude is normalized by the number of instruments. The resulting combination has a sample rate of $44100$Hz and is padded or trimmed to be exactly 4 seconds long.

@@

Partial feature dropout was performed on the feature matrix itself of input samples, the Mel spectrogram. We chose random columns and rows of the matrix to zero out. For a given matrix, each column and each row had individually a $1\%$ chance to be set to 0, which yielded an average of $1.28$  columns and $3.45$ rows being zero-ed out.

@@

To better approximate our original problem of identifying instrument and pitch, however, we performed experiments in which two instruments were selected and for input data that contained samples from one of those two instruments, both the instrument and pitch \emph{class} of the sample would be classified. The pitch class is the note name without the octave, e.g. C, D\#, G, etc. The input was a combination of two instruments drawn from a possible twelve instruments, and the classifier attempted to identify which instruments were present and for the specified instruments, say Violin and Viola, which pitch classes were present. If another instrument was present in the input combination that was not Violin or Viola, the classifier would simply identify that an instrument that was not one of the two was present. A selection of the results from this experiment are outlined in Table \ref{tab:baselines}. Since RF performed worse than SVM in every experiment, we stopped testing with it and used SVM from that point on.

@@

Our architecture is made of four convolutional layers and two fully connected layers. Each convolutional layer is followed by a BatchNorm layer, a ReLU activation layer and a $2 \times 2$  MaxPool layer with a stride of 2. The kernel size is $3 \times 3$ with a stride of 1 and a padding of 1. The number of filters are 8, 16, 32, and 32.

Following the first three convolutional layers, there is an LSTM layer which outputs 32 matrices. After the LSTM layer, there is a final convolutional layer yielding a tensor of dimensions $32\times 8 \times 21$. We flatten the outputs and feed them into a fully connected layer with Dropout, then another fully connected layer. Finally, the sigmoid function is applied to the final layer. Since each class is independent, we are able to take the sigmoid activation and use binary classification for each class. The model architecture is shown in Figure \ref{cnn_lstm}.

@@

\begin{equation}
M_N(\hat{X})_i = \left\{\begin{array}{ll}
1 \text{ if } i \in I_N(\hat{X})\\
0 \text{ otherwise}
\end{array}\right.
\label{NMax}
\end{equation}
and $I_N(\hat{X})$ is the set of indices containing the $N$ first maximums of the vector $\hat{X}$. More specifically, the function $M_N(\hat{X})$ takes as an input a vector of probabilities and outputs a vector where only the positions of the $N$ first maxima are set to 1. This new vector would be the orchestration of $N$ instruments given by the model. Thus, the function $f$ simply outputs the proportion of the estimated orchestration that matches the expected one.
For this reason, the evaluation was done by comparing the proportion of estimated orchestration samples, chosen among the samples that output the highest probability, that matched the expected orchestration.

@@

Fig. \ref{best_acc_cnn} and Fig. \ref{best_acc_resnet} show the maximum test accuracy for each model. For ResNet, the variance in accuracy is much smaller until $N$ reaches 5, at which point it becomes similar to the CNN. The results on both figures show consistency on the relative accuracy of instruments, which was for us the first step towards the validation of this method. Flute, Trombone and Trumpet yield the worst results for both models.

@@

\begin{figure}[h]
  \centering
  \begin{subfigure}{.52\textwidth}
    \centering
    \includegraphics[width=0.9\linewidth]{../ISMIR_2020/figs/Acc_vs_N_CNN.png}
    \caption{CNN with LSTM}
    \label{best_acc_cnn}
  \end{subfigure}%
  \begin{subfigure}{.52\textwidth}
    \centering
    \includegraphics[width=0.9\linewidth]{../ISMIR_2020/figs/Acc_vs_N_ResNet.png}
    \caption{ResNet}
    \label{best_acc_resnet}
  \end{subfigure}
  \caption{Plot showing per instrument accuracies for CNN with LSTM and ResNet depending on the number of samples present in the combinations}
  \end{figure}

@@

To orchestrate, a target sound is input to the model, and the 10 classes with the highest probability are extracted. These 10 classes are the instrument-pitch pairs that are most represented in the target, and can be from any combination of the 10 instruments.

@@

In order to have a better comparison, we did not allow Orchidea to use any of the advanced features: we did not apply any symbolic constraints or harmonic analysis and forced it to use all 10 instruments in each solution. This creates a more fair comparison, since our models are unable to create constrained or sparse solutions.

>>>>

\usepackage{hyperref}
\usepackage{amssymb}
\usepackage{amsmath}
\setcounter{tocdepth}{3}
\usepackage{graphicx}
\usepackage{caption}
\usepackage{subcaption}
\captionsetup{labelfont=bf, labelsep=colon,compatibility=false}
\usepackage{apacite}

% allows URLs to be broken into new lines
\def\UrlBreaks{\do\/\do-\do\&\do.\do:}

\newcommand{\keywords}[1]{\par\addvspace\baselineskip
\noindent\keywordname\enspace\ignorespaces#1}

@@

The codebase for this paper can be found at \url{https://github.com/dzluke/DeepOrchestration} and you can listen to audio examples at \url{https://dzluke.github.io/DeepOrchestration/}.

@@

\begin{enumerate}
\item We train specific models to classify the instruments present in combinations of sounds from a database of instrument notes, up to ten simultaneous instruments;
\item We then pick the best classifier and we feed into it an unknown sound to be classified;
\item Since the output of the classifier will be in the form of the probability that specific instruments-note pairs are present in the sound, we use this information to synthesize an orchestration for the target sound;
\item Finally, we evaluate the generated orchestration against state-of-the-art systems for computer-assisted orchestrations.
\end{enumerate}

@@

Two neural classifiers have been trained on the problem of recognizing instrument-pitch pairs, which has a total of 424 classes (i.e. 424 unique instrument-pitch pairs). We want to point out, however, that this is still a simplified problem compared to assisted orchestration. A complete orchestration solution would normally be made by triples of instrument-pitch-dynamics. It is difficult to frame this problem as classification, since we would need to have a very high number of classes. Moreover, for the nature of the samples we use (a typical sample is in the form \texttt{Flute-C4-pp}, as described in section \ref{sec:dataset}), each class would be represented by a single sample. For these reasons, our models have been trained on simplified versions of the original problem, and we employed a different strategy to determine dynamics. When a network is given a target sound as input, it will attempt to apply the same classification rules, outputting a vector of the probabilities of each sample being in the target sound. Since the classes only encode instrument-pitch pairs but we also need the dynamics, we quantize the probability function generated by the network into three echelons coresponding, respectively, to the dynamic levels \emph{pianissimo}, \emph{mezzoforte} and \emph{fortissimo} (see section \ref{sec:orchestration}). By taking the instruments and pitches that have the highest probability and by using quantized probabilities to retrieve dynamics, a full orchestrated solution can be created.  Fig. \ref{fig:method} illustrates the described ideas.

@@

To create the input data for training the classifiers, we used the \emph{TinySOL} database. TinySOL is a subset of the Studio On Line (SOL) database created by IRCAM \cite{Cella2020b}. TinySOL contains 1,529 samples from 12 instruments. The instruments come from different orchestral families: strings, woodwinds, and brass. Each sample is one instrument playing a single note in the \emph{ordinario} playing style, with one of three dynamics: \textit{pp}, \textit{mf}, or \textit{ff} (for example \texttt{Flute-C4-pp} or \texttt{Clarinet-D5-mf}). The instruments and ranges over which they were recorded are summarized in Tab. \ref{tab:ranges}.

For a given number of instruments $N$, each input to our model is a combination of $N$ randomly selected TinySOL samples chosen from an orchestra of ten instruments. The only constraint imposed on the random selection is that there cannot be more than three samples from the same instrument in any one combination. We do this to ensure instrumental variety in the combinations. By randomly selecting the samples, we achieve a variety of instruments, pitches, and dynamics while avoiding introducing any bias in the combinations that could result from hand-picking the samples. The selected samples are combined to be played simultaneously and the magnitude is normalized by the number of instruments. The resulting combination has a sample rate of $44100$Hz and is padded or trimmed to be exactly four seconds long.

@@

Partial feature dropout was performed on the feature matrix itself of input samples, the Mel spectrogram. We chose random columns and rows of the matrix to zero out. For a given matrix, each column and each row had individually a $1\%$ chance to be set to zero, which yielded an average of $1.28$  columns and $3.45$ rows being zero-ed out.

@@

To better approximate our original problem of identifying instrument and pitch, however, we performed experiments in which two instruments were selected and for input data that contained samples from one of those two instruments, both the instrument and pitch \emph{class} of the sample would be classified. The pitch class is the note name without the octave, e.g. C, D\#, G, etc. The input was a combination of two instruments drawn from a possible twelve instruments, and the classifier attempted to identify which instruments were present and for the specified instruments, say Violin and Viola, which pitch classes were present. If another instrument was present in the input combination that was not Violin or Viola, the classifier would simply identify that an instrument that was not one of the two was present. A selection of the results from this experiment are outlined in Tab. \ref{tab:baselines}. Since RF performed worse than SVM in every experiment, we stopped testing with it and used SVM from that point on.

@@

Our architecture is made of four convolutional layers and two fully connected layers. Each convolutional layer is followed by a BatchNorm layer, a ReLU activation layer and a $2 \times 2$  MaxPool layer with a stride of two. The kernel size is $3 \times 3$ with a stride of 1 and a padding of 1. The number of filters are eight, 16, 32, and 32.

Following the first three convolutional layers, there is an LSTM layer which outputs 32 matrices. After the LSTM layer, there is a final convolutional layer yielding a tensor of dimensions $32\times 8 \times 21$. We flatten the outputs and feed them into a fully connected layer with Dropout, then another fully connected layer. Finally, the sigmoid function is applied to the final layer. Since each class is independent, we are able to take the sigmoid activation and use binary classification for each class. The model architecture is shown in Fig. \ref{cnn_lstm}.

@@

\begin{equation}
M_N(\hat{X})_i = \left\{\begin{array}{ll}
1 \text{ if } i \in I_N(\hat{X})\\
0 \text{ otherwise}
\end{array}\right.
\label{NMax}
\end{equation}
and $I_N(\hat{X})$ is the set of indices containing the $N$ first maximums of the vector $\hat{X}$. More specifically, the function $M_N(\hat{X})$ takes as an input a vector of probabilities and outputs a vector where only the positions of the $N$ first maxima are set to one. This new vector would be the orchestration of $N$ instruments given by the model. Thus, the function $f$ simply outputs the proportion of the estimated orchestration that matches the expected one.
For this reason, the evaluation was done by comparing the proportion of estimated orchestration samples, chosen among the samples that output the highest probability, that matched the expected orchestration.

@@

Fig. \ref{best_acc} shows the maximum test accuracy for each model. For ResNet, the variance in accuracy is much smaller until $N$ reaches five, at which point it becomes similar to the CNN. The results on both figures show consistency on the relative accuracy of instruments, which was for us the first step towards the validation of this method. Flute, Trombone and Trumpet yield the worst results for both models.

@@

\begin{figure}[h]
  \centering
  \begin{subfigure}{.52\textwidth}
    \centering
    \includegraphics[width=0.9\linewidth]{../ISMIR_2020/figs/Acc_vs_N_CNN.png}
    \caption{CNN with LSTM}
    \label{best_acc_cnn}
  \end{subfigure}%
  \begin{subfigure}{.52\textwidth}
    \centering
    \includegraphics[width=0.9\linewidth]{../ISMIR_2020/figs/Acc_vs_N_ResNet.png}
    \caption{ResNet}
    \label{best_acc_resnet}
  \end{subfigure}
  \caption{Plot showing per instrument accuracies for CNN with LSTM and ResNet depending on the number of samples present in the combinations}
  \label{best_acc}
  \end{figure}

@@

To orchestrate, a target sound is input to the model, and the ten classes with the highest probability are extracted. These ten classes are the instrument-pitch pairs that are most represented in the target, and can be from any combination of the ten instruments.

@@

In order to have a better comparison, we did not allow Orchidea to use any of the advanced features: we did not apply any symbolic constraints or harmonic analysis and forced it to use all ten instruments in each solution. This creates a more fair comparison, since our models are unable to create constrained or sparse solutions.
